Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-03 article.Rmd
retooled introduction

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover publication fees, also referred to as article-processing charges (APC), authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. Yet, how and to what extent these research support activities are effective in terms of the number of supported articles and associated costs remains under debate.

One reason why studying how much is being spent on publication fees is in most cases difficult is that payments for these charges are fragmented across the budgets of funding agencies, research institutions, and libraries, or, if support is limited, are taken from personal budgets. Another key problem is that funding for open access journals using publication fees lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts, and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to variable pricing schemes include submission or page charges [@BJ_RK_2012]. Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

Previous empirical studies asking how much was being spent for publishing in open access journals confirmed that authors make use of funding or institutional support for covering publication fees. In 2010, the SOAP project carried out a comprehensive survey in which 9,645 authors from various disciplines provided insights into how they financed publication fees [@soap]. The majority of the respondents used institutional support structures or research funding. By contrast, 12 % paid publication fees individually. These findings are consistent with that of other surveys, adding that sources of funding exists mostly in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Asked about the amount being spent on publication fees, 22,977 SOAP survey respondents reported prices between 501 and 1000 € (12,9%), and between 1,001 and 3,000 € (9.9 %). However, more than half of the authors answering this question had not to pay a publication fee to get publish in an open access journal, and 14 % did not report the amount spent. While examining the indexing coverage of fully open access journals in bibliographic databases in 2009, @Walters_2011  also provided publication fees as journal characteristics. Their study reported an average publication fee of 923 $, and for Journals indexed in the Web of Science 1,171 $ on average. @Solomon_2012 calculated an similar average of 906 $ for publishing in fully open access journals. However, they found a large price variance across journals and articles. Accordingly, they suggested to cluster fully open access journals using publication fees into several groups. In this regard, while one set consisted of costly high-impact journals (2,000 $ - 4,000 $), a large number of journals charged 1,500 $ - 2,000 $. Journals with broad scope like PLOS ONE formed another segment charging between 1,000 - 1,500 $, just as mid-priced journals from commercial publishers (500 - 1,000 $), society journals (below 500 $), and low-priced journals (below 200 $).

Complementary to survey data or the consultation of journal web pages for determining spending on publication fees, some research funders and research performing institutions have recently begun to disclose their spending on publication fees as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report payments made along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. Parts of Jisc cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014.

@@

This paper focuses on how much German universities and research organization spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced through its "Open-Access Publishing" program how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. The DFG has enforced a set of criteria grantees have to comply with, leading to similar implementations for supporting open access publishing across German universities: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax). Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers. [@Fournier_2013].

@@

[^4b]: These might include article processing, impact, rejection rates, management and investment, and profit margins. See @Van_Noorden_2013 for a general discussion and @Gumpenberger_2012 and @Bj_rk_scientometrics_2015 for discussions of journal impact and quality.

>>>>

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, also referred to as article-processing charges (APC), authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. Yet, how and to what extent these research support activities are effective in terms of the number of supported articles and associated costs remains under debate.

One reason why the study of open access journals using publication fees is in most cases difficult is that total spending on publication fees is fragmented across the budgets of grant agencies, research institutions, and libraries, or, if support is limited, are taken from personal budgets. Asking 9,645 authors from various disciplines how they financed publication fees, a comprehensive survey in 2010 revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These findings are consistent with that of other studies, adding that sources of funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Another key problem for the study of open access is that funding for open access journals using publication fees lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015].  However, empirical studies examining publication fees gathered from journal websites report similar findings. While the average publication fee reported ranged between 904 $ [@Solomon_2012] and 923 $ [@Walters_2011], considerable price variances across journals and publishers were observed. Accordingly, @Solomon_2012 suggested to cluster fully open access journal using publication fees into several groups. In descending order, high-impact journals charged most, followed by biomedicine journals from commercial publishers, large multi-disciplinary journals, and mid-price ranged journals from commercial publishers covering a large spectrum of disciplines. Lower priced journals were published by academic societies and by publishers from low-income countries.

It remains unclear which other factors contribute to this varying pricing.  Generally, these might include article processing, impact, rejection rates, management and investment, and profit margins [@Van_Noorden_2013]. While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts, and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to variable pricing schemes include submission or page charges [@BJ_RK_2012].

Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

To address the problems of fragmented spending on publication fees and in-transparency about what was paid, some European research funders and research performing institutions have recently begun to disclose their spending on publication fees as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report payments made along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates [@woodward_2014]. Parts of Jisc cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014.

@@

This paper focuses on how much German universities and research organization spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program  in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax). Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers [@Fournier_2013].

@@
